---
title: "exercise-05"
format: html
---

###################### Challenge 1

## Step 1: Load Data

```{r}
#| message: false
#| warning: false
library(tidyverse)

d <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/IMDB-movies.csv")

head(d)
nrow(d)
glimpse(d)
summary(d)
```

## Step 2: Filter and Add Decade

```{r}
d <- d %>%
  filter(startYear >= 1920 & startYear <= 1979 & runtimeMinutes >= 60 & runtimeMinutes <= 180) %>%
  mutate(decade = case_when(
    startYear >= 1920 & startYear <= 1929 ~ "20s",
    startYear >= 1930 & startYear <= 1939 ~ "30s",
    startYear >= 1940 & startYear <= 1949 ~ "40s",
    startYear >= 1950 & startYear <= 1959 ~ "50s",
    startYear >= 1960 & startYear <= 1969 ~ "60s",
    startYear >= 1970 & startYear <= 1979 ~ "70s"
  ))

table(d$decade)
```

## Step 3: Histograms by Decade

```{r}
ggplot(data = d, aes(x = runtimeMinutes)) +
  geom_histogram() +
  facet_wrap(~decade)
```

## Step 4: Mean and SD by Decade

```{r}
results <- d %>%
  group_by(decade) %>%
  summarize(mean = mean(runtimeMinutes, na.rm = TRUE),
            sd = sd(runtimeMinutes, na.rm = TRUE))

results
```

## Step 5: Sample Statistics

```{r}
sample1 <- d %>%
  group_by(decade) %>%
  slice_sample(n = 100, replace = FALSE) %>%
  summarize(sample_mean = mean(runtimeMinutes),
            sample_sd = sd(runtimeMinutes))

sample1
```

## Step 6: Standard Error

```{r}
sample1 <- sample1 %>%
  mutate(n = 100) %>%
  mutate(se = sample_sd / sqrt(n))

sample1
```

## Step 7: Compare Sample to Population

```{r}
comparison <- results %>%
  rename(pop_mean = mean, pop_sd = sd) %>%
  left_join(sample1, by = "decade")

comparison <- comparison %>%
  mutate(expected_se = pop_sd / sqrt(100)) %>%
  mutate(mean_diff = sample_mean - pop_mean) %>%
  mutate(se_diff = se - expected_se)

comparison
```

## Step 8: Sampling Distribution

```{r}
#| message: false
#| warning: false
sampling_dist <- data.frame()

for(i in 1:1000) {
  sample_i <- d %>%
    group_by(decade) %>%
    sample_n(size = 100, replace = FALSE) %>%
    summarize(sample_mean = mean(runtimeMinutes),
              sample_sd = sd(runtimeMinutes)) %>%
    mutate(replicate = i)

  sampling_dist <- rbind(sampling_dist, sample_i)
}

head(sampling_dist)
dim(sampling_dist)
```

```{r}
sampling_dist_summary <- sampling_dist %>%
  group_by(decade) %>%
  summarize(mean_of_means = mean(sample_mean),
            sd_of_means = sd(sample_mean))

sampling_dist_summary
comparison
```

## Step 9: Sampling Distribution Plots and SE Comparison

The sampling distributions are bell-shaped, demonstrating the Central Limit Theorem. Sample means closely approximate population means as expected.

```{r}
sampling_dist_summary <- sampling_dist %>%
  group_by(decade) %>%
  summarize(mean_of_means = mean(sample_mean),
            sd_of_means = sd(sample_mean))

sampling_dist_summary

ggplot(data = sampling_dist, aes(x = sample_mean)) +
  geom_histogram(bins = 30) +
  facet_wrap(~decade) +
  labs(title = "Sampling Distribution of Sample Means by Decade",
       x = "Sample Mean Runtime (Minutes)",
       y = "Frequency")
```

```{r}
se_comparison <- comparison %>%
  left_join(sampling_dist_summary, by = "decade") %>%
  select(decade, se, expected_se, sd_of_means) %>%
  rename(se_from_sample = se,
         se_from_population = expected_se,
         se_from_sampling_dist = sd_of_means)

se_comparison
```

---

########################################################### Challenge 2

## Step 1: Load Zombie Data

```{r}
#| message: false
#| warning: false
z <- read_csv("https://raw.githubusercontent.com/difiore/ada-datasets/main/zombies.csv")
head(z)
glimpse(z)
dim(z)
```

## Step 2: Population Means and SDs

```{r}
mean_height <- mean(z$height)
sd_height <- sqrt(sum((z$height - mean_height)^2) / nrow(z))

mean_weight <- mean(z$weight)
sd_weight <- sqrt(sum((z$weight - mean_weight)^2) / nrow(z))

mean_age <- mean(z$age)
sd_age <- sqrt(sum((z$age - mean_age)^2) / nrow(z))

mean_zombies <- mean(z$zombies_killed)
sd_zombies <- sqrt(sum((z$zombies_killed - mean_zombies)^2) / nrow(z))

mean_education <- mean(z$years_of_education)
sd_education <- sqrt(sum((z$years_of_education - mean_education)^2) / nrow(z))

mean_height; sd_height
mean_weight; sd_weight
mean_age; sd_age
mean_zombies; sd_zombies
mean_education; sd_education
```

## Step 3: Boxplots by Gender

```{r}
z %>%
  select(gender, height, weight, age, zombies_killed, years_of_education) %>%
  pivot_longer(cols = -gender, names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = gender, y = value, fill = gender)) +
  geom_boxplot() +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Distribution of Variables by Gender",
       x = "Gender", y = "Value")
```

## Step 4: Scatterplots

Height vs Age: The scatterplot shows a weak positive correlation between height and age for both genders. As age increases, height tends to increase slightly, possibly because younger individuals are still growing.

```{r}
ggplot(data = z, aes(x = age, y = height)) +
  geom_point(aes(color = gender))
```

Weight vs Age: The scatterplot shows no clear relationship between weight and age. The points are scattered randomly with no obvious pattern, indicating little to no correlation.

```{r}
ggplot(data = z, aes(x = age, y = weight)) +
  geom_point(aes(color = gender))
```

## Step 5: Normality Check

Height, weight, and age all look pretty normal. The histograms are bell-shaped and symmetric, and in the Q-Q plots the points follow the diagonal line pretty closely. Zombies killed and years of education don't look normal — they follow a Poisson distribution since they count discrete events.

```{r}
z %>%
  select(height, weight, age, zombies_killed, years_of_education) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30) +
  facet_wrap(~variable, scales = "free")
```

```{r}
z %>%
  select(height, weight, age, zombies_killed, years_of_education) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(sample = value)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~variable, scales = "free")
```

## Step 6: Sample of 50 and Confidence Intervals

```{r}
sample_50 <- z %>% sample_n(size = 50, replace = FALSE)

mean_height_sample <- mean(sample_50$height)
mean_weight_sample <- mean(sample_50$weight)
mean_age_sample <- mean(sample_50$age)
mean_zombies_sample <- mean(sample_50$zombies_killed)
mean_education_sample <- mean(sample_50$years_of_education)

sd_height_sample <- sd(sample_50$height)
sd_weight_sample <- sd(sample_50$weight)
sd_age_sample <- sd(sample_50$age)
sd_zombies_sample <- sd(sample_50$zombies_killed)
sd_education_sample <- sd(sample_50$years_of_education)

n <- 50
se_height <- sd_height_sample / sqrt(n)
se_weight <- sd_weight_sample / sqrt(n)
se_age <- sd_age_sample / sqrt(n)
se_zombies <- sd_zombies_sample / sqrt(n)
se_education <- sd_education_sample / sqrt(n)

critical_value <- qt(0.975, df = 49)

ci_lower_height <- mean_height_sample - critical_value * se_height
ci_upper_height <- mean_height_sample + critical_value * se_height

ci_lower_weight <- mean_weight_sample - critical_value * se_weight
ci_upper_weight <- mean_weight_sample + critical_value * se_weight

ci_lower_age <- mean_age_sample - critical_value * se_age
ci_upper_age <- mean_age_sample + critical_value * se_age

ci_lower_zombies <- mean_zombies_sample - critical_value * se_zombies
ci_upper_zombies <- mean_zombies_sample + critical_value * se_zombies

ci_lower_education <- mean_education_sample - critical_value * se_education
ci_upper_education <- mean_education_sample + critical_value * se_education

mean_height_sample; se_height; ci_lower_height; ci_upper_height
mean_weight_sample; se_weight; ci_lower_weight; ci_upper_weight
mean_age_sample; se_age; ci_lower_age; ci_upper_age
mean_zombies_sample; se_zombies; ci_lower_zombies; ci_upper_zombies
mean_education_sample; se_education; ci_lower_education; ci_upper_education
```

## Step 7: Sampling Distribution of 200 Samples

```{r}
set.seed(456)
sampling_dist_200 <- replicate(199, {
  z %>%
    slice_sample(n = 50) %>%
    summarize(
      mean_height = mean(height),
      mean_weight = mean(weight),
      mean_age = mean(age),
      mean_zombies = mean(zombies_killed),
      mean_education = mean(years_of_education)
    )
}, simplify = FALSE) %>%
  bind_rows()

first_sample_means <- sample_50 %>%
  summarize(
    mean_height = mean(height),
    mean_weight = mean(weight),
    mean_age = mean(age),
    mean_zombies = mean(zombies_killed),
    mean_education = mean(years_of_education)
  )

sampling_dist_200 <- bind_rows(first_sample_means, sampling_dist_200)

sampling_summary <- sampling_dist_200 %>%
  summarize(
    mean_of_height_means = mean(mean_height),
    sd_of_height_means = sd(mean_height),
    mean_of_weight_means = mean(mean_weight),
    sd_of_weight_means = sd(mean_weight),
    mean_of_age_means = mean(mean_age),
    sd_of_age_means = sd(mean_age),
    mean_of_zombies_means = mean(mean_zombies),
    sd_of_zombies_means = sd(mean_zombies),
    mean_of_education_means = mean(mean_education),
    sd_of_education_means = sd(mean_education)
  )

sampling_summary

comparison_se <- tibble(
  variable = c("height", "weight", "age", "zombies_killed", "years_of_education"),
  se_from_sample = c(se_height, se_weight, se_age, se_zombies, se_education),
  sd_of_sampling_dist = c(
    sampling_summary$sd_of_height_means,
    sampling_summary$sd_of_weight_means,
    sampling_summary$sd_of_age_means,
    sampling_summary$sd_of_zombies_means,
    sampling_summary$sd_of_education_means
  )
)

comparison_se
```

The SD of the sampling distribution is pretty close to the SE from the first sample for all variables. This shows that the SE formula works well to estimate how much sample means will vary. They're not exactly the same because of random variation, but they're close enough to show the theory works in practice.

## Step 8: Histograms and Q-Q Plots of Sampling Distributions

All the sampling distributions look pretty normal and bell-shaped. Notably, zombies_killed and years_of_education look normal even though the original data was Poisson distributed. This demonstrates the Central Limit Theorem — when you take means of samples large enough (n=50), the distribution of those means becomes normal regardless of the original distribution.

```{r}
sampling_dist_200 %>%
  pivot_longer(everything(), names_to = "variable", values_to = "sample_mean") %>%
  ggplot(aes(x = sample_mean)) +
  geom_histogram(bins = 25) +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Sampling Distributions of Sample Means",
       x = "Sample Mean", y = "Frequency")
```

```{r}
sampling_dist_200 %>%
  pivot_longer(everything(), names_to = "variable", values_to = "sample_mean") %>%
  ggplot(aes(sample = sample_mean)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Q-Q Plots of Sampling Distributions")
```

## Step 9: 95% CIs from Sampling Distribution

```{r}
ci_from_sampling <- tibble(
  variable = c("height", "weight", "age", "zombies_killed", "years_of_education"),
  ci_lower = c(
    quantile(sampling_dist_200$mean_height, 0.025),
    quantile(sampling_dist_200$mean_weight, 0.025),
    quantile(sampling_dist_200$mean_age, 0.025),
    quantile(sampling_dist_200$mean_zombies, 0.025),
    quantile(sampling_dist_200$mean_education, 0.025)
  ),
  ci_upper = c(
    quantile(sampling_dist_200$mean_height, 0.975),
    quantile(sampling_dist_200$mean_weight, 0.975),
    quantile(sampling_dist_200$mean_age, 0.975),
    quantile(sampling_dist_200$mean_zombies, 0.975),
    quantile(sampling_dist_200$mean_education, 0.975)
  )
)

ci_from_sampling

ci_comparison <- tibble(
  variable = c("height", "weight", "age", "zombies_killed", "years_of_education"),
  ci_lower_sample = c(ci_lower_height, ci_lower_weight, ci_lower_age, ci_lower_zombies, ci_lower_education),
  ci_upper_sample = c(ci_upper_height, ci_upper_weight, ci_upper_age, ci_upper_zombies, ci_upper_education),
  ci_lower_quantile = ci_from_sampling$ci_lower,
  ci_upper_quantile = ci_from_sampling$ci_upper,
  pop_mean = c(mean_height, mean_weight, mean_age, mean_zombies, mean_education)
)

ci_comparison
```

## Step 10: Bootstrap 95% CIs

```{r}
set.seed(123)

boot_height <- numeric(1000)
boot_weight <- numeric(1000)
boot_age <- numeric(1000)
boot_zombies <- numeric(1000)
boot_education <- numeric(1000)

for(i in 1:1000) {
  boot_sample <- sample_50[sample(nrow(sample_50), nrow(sample_50), replace = TRUE), ]
  boot_height[i] <- mean(boot_sample$height)
  boot_weight[i] <- mean(boot_sample$weight)
  boot_age[i] <- mean(boot_sample$age)
  boot_zombies[i] <- mean(boot_sample$zombies_killed)
  boot_education[i] <- mean(boot_sample$years_of_education)
}

ci_bootstrap <- tibble(
  variable = c("height", "weight", "age", "zombies_killed", "years_of_education"),
  ci_lower_boot = c(
    quantile(boot_height, 0.025),
    quantile(boot_weight, 0.025),
    quantile(boot_age, 0.025),
    quantile(boot_zombies, 0.025),
    quantile(boot_education, 0.025)
  ),
  ci_upper_boot = c(
    quantile(boot_height, 0.975),
    quantile(boot_weight, 0.975),
    quantile(boot_age, 0.975),
    quantile(boot_zombies, 0.975),
    quantile(boot_education, 0.975)
  )
)

ci_bootstrap

ci_all <- ci_comparison %>%
  left_join(ci_bootstrap, by = "variable")

ci_all
```

The bootstrap CIs from Step 10 are very similar to both CI methods from Step 9. Compared to the single-sample CIs, the bootstrap intervals are nearly identical since both methods are drawing from the same original sample of 50. Compared to the sampling distribution CIs from Step 9, the bootstrap CIs are slightly wider for some variables, which makes sense since the sampling distribution was built from 200 draws from the full population while bootstrapping only resamples from one sample. All three methods capture the population mean within their bounds, suggesting that bootstrapping is a reliable way to estimate confidence intervals even without access to the full population.
````

